PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2022-08-16 main.tex
More edits

Motivated by the dependence of accretion modes on $\alpha$,
we next investigate whether this highly efficient, non-oligarchic
growth mode should be expected to operate in a typical planet-forming
disk. Given that N-body simulations of short-period terrestrial planet formation 
typically begin with a chain of neatly-spaced, isolation mass 
(see \citet{kokubo00} eq. 20) protoplanets, it is pertinent to determine 
whether the high $\alpha$ growth mode we revealed
in the previous section invalidates this choice of initial conditions.

 Given the dearth of short-period terrestrial planets observed around M stars (e.g. TRAPPIST-1), we chose to model a series of wide planetesimal disks, which span from 1 to 100 days in orbital period orbiting a late-type M star of mass 0.08 $M_{\odot}$. For a population of planetesimals with a bulk density of 3 g cm$^{-3}$, this orbital period range corresponds to $\alpha \in (0.7, 0.05)$. By simultaneously modeling a broad range of orbital periods, we can determine the critical value of $\alpha$ that divides these two modes of accretion, and also explore how the oligarchic/non-oligarchic accretion boundary affects the resulting distribution of protoplanets.

Four wide disk simulations are run in total (see table \ref{tab:sims}). In each case, the solid surface density follows a power law profile
\begin{equation}
	\Sigma(r) = \left( \frac{M_{*}}{M_{\odot}} \right) A \Sigma_{\mathrm{MMSN}} \left( \frac{r}{1 \mathrm{AU}} \right)^{-\delta},
\end{equation}
where $M_{*}$ is the mass of the central star, $\Sigma_{MMSN} = 10$ g cm$^{-2}$ is surface density of the minimum-mass solar nebula \citep{hayashi81} at 1 AU, and $A$ is an enhancement factor. In the first case (fdHi), we model a disk that follows a MMSN power law slope, with the overall normalization enhanced by a factor of 100. This choice of parameters for the solid surface density profile appears necessary in order to reproduce many observed short period terrestrial worlds in-situ \citep{hansen12}. Additionally, we vary the power law index (fdHiShallow, fdHiSteep) and overall normalization (fdLo) of the solids profile.

@@

In addition to the mutual gravitational forces, a Stokes drag force from to the gas disk is applied to each particle, following the prescription described in section 2.2.1 of \citet{morishima10}. The surface density and temperature profile are identical to those used to construct the initial conditions and the cylindrical coordinate system is centered on the host star.

@@

In the case of the fdHi simulation, there are nearly 1 million
particles, whose orbital periods vary by two orders of magnitude. Due
to the very short interaction timescales near the inner edge of the
disk, a fixed timestep size would required a prohibitively large
number of steps to follow planetesimal growth throughout the entire
disk. For this reason, we use a multi-tiered timestepping scheme, in
which particles are placed onto the nearest power of two timestep
based on their most recently calculated gravitational
acceleration. (This scheme is used on almost all works using ChaNGa,
and is common among large-scale simulation codes.)

This more efficient scheme, however, introduces two additional issues
related to particles interacting on different timesteps. First,
momentum is not completely conserved when bodies switch timesteps. The
error introduced becomes particularly severe for a particle on an
eccentric Keplerian orbit, whose perihelion and aphelion distances
straddle a timestep boundary. For a large collection of particles,
this problem manifests itself as the development of a V-shaped gap in the semimajor axis-eccentricity plane, which is centered on the boundary itself. To correct this problem, we introduce a second timestepping criterion, which is based on the gravitational acceleration on the particle by the central star at perihelion. Only in the case of a close encounter with another planetesimal (in which the acceleration is no longer dominated by the star) is the timestep allowed to reduce based on the original criterion.

A second issue is introduced when two particles on different timesteps undergo a collision. As in the previous case, momentum is not completely conserved because the most recent `kick' steps did not happen simultaneously for these bodies. Early in the simulation, we find that runaway growth tends to trigger first at the timestep boundaries. Unfortunately, this issue carries itself forward through the embryo formation phase and protoplanets tend to form at the boundaries. To correct this issue, we ignore collisions between bodies on different timesteps early in the simulation. We find that enabling multi-timestep collisions after the maximum mass grows by a factor of 10 prevents any artifacts from forming at the timestep boundaries, while also minimizing the number of `skipped' collisions\footnote{In the case of fdHi, only about 20 collisions out of an eventual 900,000 are ignored.}.

@@

\begin{table}
\caption{Final Properties of Full Disk Simulations}
%TRQ: the mass numbers in this table look wrong.
\begin{tabular}{lllll} \hline \hline
Name     & $M_{PP}^a$ & $T_{int}^b$ & $T_{int 1}^c$ &  \\ \hline
fdHi                 & 100                                                         & 1.5                              & 4  &  \\
fdHiShallow     & 100                                                        & 0.5                              & 4  &  \\
fdHiSteep       & 100                                                         & 2.5                              & 4 &  \\
fdLo                & 1                                                             & 1.5                              & 4 &  \\ \hline
\end{tabular}
\label{tab:sim_properties}
\begin{flushleft}
\textsc{Note} \\ {$^a$ Maximum protoplanet mass [$M_{earth}$] \\
		       $^b$ Integration time [yr] \\
		       $^c$ Rescaled integration time [yr]} \\
\end{flushleft}
\end{table}

In all four simulations, the timescales for embryo formation depend on
the chosen surface density profile, along with the particular location
in the disk. Protoplanets form first near the inner edge of the disk,
where the dynamical timescales are short. Growth proceeds in an
inside-out fashion, with the outermost regions of the disk completing
the protoplanet assembly phase last. This fact is not typically
accounted for in planet formation simulations, and appears to be an
important component to forming realistic solar system analogs
\citep{clement20}. As with the narrow annulus simulations, we stop the
integration once the masses of protoplanets in the outermost region of
the disk reach a steady value. In table \ref{tab:sim_properties}, we
summarize the outcomes of the four ``full disk'' cases.

We show the final state of the ``fiHi'' simulation in figure \ref{fig:fulldisk_e_m}. In the top panel, the initial (contours) and final (points) state of the simulation is shown in orbital period and eccentricity. The size of the points indicates the relative mass of the bodies. In the bottom panel, the mass of the largest bodies is shown as a function of orbital period. The solid curves indicate the isolation mass,
\begin{equation}\label{eq:iso}
	M_{iso} = \left[ \frac{\left( 2 \pi a^2 \Sigma \tilde{b} \right)^3}{3 M_{*}} \right]^{1/2},
\end{equation}
where $\tilde{b}$ is the size of the feeding zone in units
of Hill radii. For a body on a circular orbit $\tilde{b} = 2
\sqrt{3}$, while a protoplanet formed by oligarchic growth typically
has $\tilde{b} = 10$ \citep{kokubo98}.
%TRQ there is no orange line in the figure.
The isolation mass is plotted in blue and orange for these two values of $\tilde{b}$, respectively.

A qualitative shift in the protoplanet and planetesimal distribution is visible at $\sim$ 60 days. Interior to this location, there are very few remaining planetesimals and the embryos formed are noticeably more massive. Protoplanets in this region appear to closely follow the isolation mass curve for $\tilde{b} = 10$. Exterior to the boundary, the residual planetesimal population is much more numerous, and protoplanets more closely follow the $\tilde{b} = 2 \sqrt{3}$ curve. This tentatively suggests that the transition between the low $\alpha$ and high $\alpha$ accretion modes seen in section \ref{sec:narrow} happens near this location.

% Also mention that no eccentricity difference between large and small alpha regions, as we saw in the narrow annulus simulations

In figure (removed), we show $\alpha$ (left axis) and the fraction of remaining planetesimals (right axis) as a function of orbital period. Here, the fraction of remaining planetesimals is measured by dividing the disk into 15 semimajor axis bins and using the profile function in {\sc pynbody} to measure the total solid surface density $\Sigma$ in each of those bins, and also the planetesimal surface density $\sigma$. Here, a planetesimal is defined to any body with $m < 10 m_{pl}$. Exterior to 60 days in orbital period, $\sigma/\Sigma$ sharply increases, indicating that the highly efficient accretion mode seen previously does not operate here. As can be seen from the plot, the accretion boundary roughly corresponds to $\alpha = 0.1$.
% This figure could be turned into a few sentences in the text.

@@

To ensure that the boundary seen near 60 days in orbital period is not
simply a transient product of the inside out growth throughout the
disk, we examine the time evolution of $\sigma/\Sigma$ at multiple
locations. In figure \ref{fig:pl_frac_time}, the value of
$\sigma/\Sigma$ is plotted as a function of time in 10 orbital period
bins.
%TRQ: width of the bins?
Bins interior to 60 days are indicated with a dashed line, while those exterior are indicated with a solid line. The planetesimal surface density fraction has decelerated and is approaching an asymptotic value in all cases. In the inner disk, this is driven by the near-total depletion of the planetesimal reservoir. In the outer disk, dynamical friction between the embryos and planetesimals throttles subsequent accretion.

% Since we are using f=6, do we want to stick to plotting things in terms of timestep number, and then provide rescaled time estimates at the end?

@@

In figure \ref{fig:surfden_iso}, we compare the masses of the resulting protoplanets and planetesimals in all four simulations to each other. Here, the masses are scaled by the isolation mass at the given orbital distance (see equation \ref{eq:iso}). In all four cases, the masses follow a similar trend: interior to about 60 days in orbital period, protoplanets grow to a factor of a few larger than $M_{iso}$. Exterior to this point, the protoplanet masses roughly stay near the isolation mass. Despite the drastically different solid surface density profiles used, the masses of the largest bodies in all four cases follow a similar trend with orbital period. This indicates that the highly efficient accretion mode seen at shorter orbital periods is dependent on the sizes of the individual bodies themselves, rather than how they are distributed. The transition visible near 60 days in all cases also supports our conclusion that planetesimal accretion is largely complete everywhere in the disk.
% fdLo needs to run a bit longer

>>>>

Motivated by the dependence of accretion modes on $\alpha$,
we next investigate whether this highly efficient, non-oligarchic
growth should be expected to operate near the innermost regions of a typical planet-forming
disk. Given that N-body simulations of short-period terrestrial planet formation 
typically begin with a chain of neatly-spaced, isolation mass 
(see \citet{kokubo00} eq. 20) protoplanets, it is pertinent to determine 
whether the high $\alpha$ growth mode we revealed
in the previous section invalidates this choice of initial conditions.

Given the dearth of short-period terrestrial planets observed around M stars (e.g. TRAPPIST-1), we chose to model the evolution of a series of wide planetesimal disks, which span from 1 to 100 days in orbital period, orbiting a late-type M star of mass 0.08 $M_{\odot}$. For a population of planetesimals with a bulk density of 3 g cm$^{-3}$, this orbital period range corresponds to $\alpha \in (0.7, 0.05)$. By simultaneously modeling a broad range of orbital periods, we can determine the critical value of $\alpha$ that divides these two modes of accretion, and also explore how the oligarchic/non-oligarchic accretion boundary affects the resulting distribution of protoplanets.

Four wide-disk simulations are run in total (see table \ref{tab:sims}). In each case, the solid surface density follows a power law profile
\begin{equation}
	\Sigma(r) = \left( \frac{M_{*}}{M_{\odot}} \right) A \Sigma_{\mathrm{MMSN}} \left( \frac{r}{1 \mathrm{AU}} \right)^{-\delta},
\end{equation}
where $M_{*}$ is the mass of the central star, $\Sigma_{MMSN} = 10$ g cm$^{-2}$ is surface density of the minimum-mass solar nebula \citep{hayashi81} at 1 AU, and $A$ is an enhancement factor. In the first case (fdHi), we model a disk that follows a MMSN power law slope, with the overall normalization enhanced by a factor of 100. This choice of normalization for the solid surface density profile appears necessary in order to reproduce many observed short period terrestrial worlds in-situ \citep{hansen12}. Additionally, we vary the power law index (fdHiShallow, fdHiSteep) and overall normalization (fdLo) of $\Sigma(r)$.

@@

In addition to the mutual gravitational forces, a Stokes drag force from to the gas disk is applied to each particle, following the prescription described in section 2.2.1 of \citet{morishima10}.

@@

In the case of the fdHi simulation, there are nearly 1 million
particles, whose orbital periods vary by two orders of magnitude. Because the
interaction timescales near the inner edge of the
disk are exceedingly short, a fixed timestep size would required a prohibitively large
number of steps to follow planetesimal growth throughout the entire
disk. For this reason, we use a multi-tiered timestepping scheme, in
which particles are placed onto the nearest power of two timestep
based on their most recently calculated gravitational
acceleration. \footnote{This scheme is used on almost all works using ChaNGa,
and is common among large-scale simulation codes.}

This more efficient scheme introduces two issues, however. Firstly,
momentum is not completely conserved when bodies switch timesteps. The
error introduced becomes particularly severe for a particle on an
eccentric Keplerian orbit, whose perihelion and aphelion distances
straddle a timestep boundary. For a large collection of particles,
this problem manifests itself as the development of a V-shaped gap in the semimajor axis-eccentricity plane, which is centered on the boundary itself. To correct this problem, we introduce a slightly modified timestepping criterion, which is based on the expected gravitational acceleration of the particle at pericenter. Only in the case of a close encounter with another planetesimal (in which the acceleration is no longer dominated by the star) is the timestep allowed to reduce based on the original criterion.

A second issue is introduced when two particles on different timesteps undergo a collision. As in the previous case, momentum is not completely conserved because the most recent `kick' steps did not happen simultaneously for these bodies. Early in the simulation, we find that runaway growth tends to initiate at the timestep boundaries. This issue carries itself forward through the embryo formation phase, and protoplanets tend to form at the boundaries. To correct this issue, we ignore collisions between bodies on different timesteps early in the simulation. We find that preventing multi-timestep collisions until after the maximum mass grows by a factor of 10 prevents any artifacts from developing at the timestep boundaries, while also minimizing the number of `skipped' collisions\footnote{In the case of fdHi, only about 20 collisions out of an eventual 900,000 are ignored.}.

@@

\begin{table}
\caption{Final Properties of Full Disk Simulations}
\begin{tabular}{lllll} \hline \hline
Name     & $M_{PP}^a$ & $T_{int}^b$ & $T_{int 1}^c$ &  \\ \hline
fdHi                 &  1.00                                                         & 456                              & 16,377  &  \\
fdHiShallow    &  1.19                                                        & 456                              & 16,377  &  \\
fdHiSteep       &  1.08                                                         & 456                              & 16,377 &  \\
fdLo                &  1.77 $\times 10^{-3}$                                  & 3,713                              & 133,651 &  \\ \hline
\end{tabular}
\label{tab:sim_properties}
\begin{flushleft}
\textsc{Note} \\ {$^a$ Maximum protoplanet mass [$M_{earth}$] \\
		       $^b$ Integration time [yr] \\
		       $^c$ Rescaled integration time [yr]} \\
\end{flushleft}
\end{table}

The timescales for embryo formation depend on
the chosen surface density profile, along with the local orbital
timescale. Protoplanets form first at the inner edge of the disk,
where the dynamical timescales are short. Growth proceeds in an
inside-out fashion, with the outermost regions of the disk completing
the protoplanet assembly phase last. This radial timescale dependence is not typically
accounted for in planet formation simulations\footnote{Instead, the innermost protoplanets 
patiently `wait' for this phase of evolution to complete everywhere in the disk}, and appears to be an
important component to forming realistic solar system analogs
\citep{clement20}. As with the narrow annulus simulations, we stop the
integration once the masses of protoplanets in the outermost region of
the disk reach a steady value. In table \ref{tab:sim_properties}, we
summarize the outcomes of the four ``full disk'' cases.

We show the final state of the ``fdHi'' simulation in figure \ref{fig:fulldisk_e_m}. In the top panel,
the initial (contours) and final (points) state of the simulation is shown in the orbital period-eccentricity plane. The size of the points indicates the relative mass of the bodies. In the bottom panel, the mass of the largest bodies is shown as a function of orbital period. The solid curves indicate the isolation mass,
\begin{equation}\label{eq:iso}
	M_{iso} = \left[ \frac{\left( 2 \pi a^2 \Sigma \tilde{b} \right)^3}{3 M_{*}} \right]^{1/2},
\end{equation}
where $\tilde{b}$ is the size of the feeding zone in units
of Hill radii. For two bodies on circular, non-inclined orbits,  $\tilde{b} = 2
\sqrt{3}$ is the smallest orbital separation that produces a non-negative Jacobi energy
and permits a close encounter (see \citet{naka88}).
The isolation mass is plotted in blue and orange for these two values of $\tilde{b}$, respectively.

A qualitative shift in the protoplanet and planetesimal distribution is visible at $\sim$ 60 days. Interior to this location, there are very few remaining planetesimals and the embryos formed are noticeably more massive. Protoplanets in this region appear to closely follow the isolation mass curve for $\tilde{b} = 10$. Exterior to the boundary, the residual planetesimal population is much more pristine, and protoplanets more closely follow the $\tilde{b} = 2 \sqrt{3}$ curve. This suggests that the transition between the low $\alpha$ and high $\alpha$ accretion modes seen in section \ref{sec:narrow} happens near this location.

In section \ref{sec:narrow}, we postulated that the increased importance of inelastic damping in the inner, non-oligarchic growth region of the disk should lower the overall eccentricity of the protoplanets there. This behavior is not immediately apparent in the top panel of figure \ref{fig:fulldisk_e_m}. There are, however, a couple of factors in the wide disk simulations that could make this extra dynamical cooling mechanism difficult to see. Firstly, the initial eccentricity distributions of the inner and outer disk are different because of the dependence of the viscous stirring and gas drag timescales on orbital period. Additionally, the protoplanet formation timescales for the inner and outer disk are vastly different, rendering a comparison between these regions at the same moment in time inappropriate.

%In figure (removed), we show $\alpha$ (left axis) and the fraction of remaining planetesimals (right axis) as a function of orbital period. Here, the fraction of remaining planetesimals is measured by dividing the disk into 15 semimajor axis bins and using the profile function in {\sc pynbody} to measure the total solid surface density $\Sigma$ in each of those bins, and also the planetesimal surface density $\sigma$. Here, a planetesimal is defined to any body with $m < 10 m_{pl}$. Exterior to 60 days in orbital period, $\sigma/\Sigma$ sharply increases, indicating that the highly efficient accretion mode seen previously does not operate here. As can be seen from the plot, the accretion boundary roughly corresponds to $\alpha = 0.1$.
% This figure could be turned into a few sentences in the text.

@@

To ensure that the boundary seen near 60 days in orbital period is not
simply a transient product of the inside-out growth throughout the
disk, we examine the time evolution of $\sigma/\Sigma$ at multiple
locations. In figure \ref{fig:pl_frac_time}, the value of
$\sigma/\Sigma$ is plotted as a function of time in 10 orbital period
bins, each with a width of 10 days.
Bins interior to 60 days are indicated with a dashed line, while those exterior are indicated with a solid line. In all radial bins, the planetesimal surface density is approaching an asymptotic value. In the inner disk, this value asymptotes to zero as the planetesimal population completely depletes. In the outer disk, dynamical friction between the embryos and planetesimals eventually throttles subsequent accretion and leaves $\sim$ 10 to 20 percent of the mass surface density as planetesimals.

@@

In figure \ref{fig:surfden_iso}, we compare the masses of the resulting protoplanets and planetesimals in all four simulations to each other. Here, the masses are scaled by the isolation mass at the given orbital distance (see equation \ref{eq:iso}). In all four cases, the masses follow a similar trend: interior to about 60 days in orbital period, protoplanets grow to a factor of a few larger than $M_{iso}$. Exterior to this point, the protoplanet masses roughly stay near the isolation mass. Despite the drastically different solid surface density profiles used, the masses of the largest bodies in all four cases follow a similar trend with orbital period. This indicates that the highly efficient accretion mode seen at shorter orbital periods is dependent on the sizes of the individual bodies themselves, rather than how they are distributed. The transition visible near 60 days in all cases also supports our conclusion that planetesimal accretion is largely complete everywhere in the disk.
